---
header-includes:
- \usepackage{xcolor}
- \usepackage{color} 
- \usepackage{fancyhdr,color}
- \usepackage{lipsum}
- \fancyfoot[CE] {\thepage}
title: "Assignment 3"
subtitle: "Applied Forecasting in Complex Systems 2025"
author: Bugra Sipahioglu (14318334), Despoina Delipalla (16324897), Jeremi Wasilewski (14271028)
date: "University of Amsterdam \n &nbsp;  \n December 23, 2025"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = 'center',   # center plots
  #fig.width = 5,          # width in inches (smaller than default 6)
  #fig.height = 2,         # height in inches (smaller than default 4)
  fig.pos = 'H',          # keep figures in place
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  dev.args = list(pointsize = 11)
)
options(digits = 3, width = 60)

#install.packages("readr")
#install.packages("tsibble")

library(readr)
library(dplyr)
library(lubridate)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(tsibble)
library(ggplot2)
library(fpp3)
library(fable)
library(feasts)
```

# Exploratory Data Analysis

#### Load the Data

```{r}
calendar <- read_csv("../data/calendar_afcs2025.csv")  # date, event_type, week_id
sell_prices <- read_csv("../data/sell_prices_afcs2025.csv")  # week_id, item_id, price
sales_data  <- read_csv("../data/sales_train_validation_afcs2025.csv")  # item_id, product_type, items_sold
     
```

#### Calendar

```{r}
# Transform date format to date and ensure that dates are in chronological order
calendar <- calendar %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y"))%>%
 arrange(date)


# Add column day_id
calendar <- calendar %>%   
  mutate(day_id = paste0("id_", row_number()))
```

```{r}
# rows and columns
no_of_rows = nrow(calendar)
no_of_cols = ncol(calendar)
col_names = colnames(calendar)

# Dates
range = range(calendar$date)

# NA values
rows_na = colSums(is.na(calendar))

# Duplicates
dupls = anyDuplicated(calendar)


print("Number of rows") 
print(no_of_rows)
print("-----------------------")
print("Number of columns")
print(no_of_cols)
print("-----------------------")
print("Column names:")
print(col_names)
print("-----------------------")
print("Period covered")
print(range)
#TODO: it's not defined anywhere above this code. 
print("-----------------------")
print("Gaps in dates")
#print(gaps_in_dates) 
print("-----------------------")
print("Rows with na values")
print(rows_na)
print("-----------------------")
print("Number of duplicates")
print(dupls)
```

Colendard contains information about the dates the products are sold.
Calendar contains columns:

-   date: The date in a "y-m-d" format.

-   wm_yr_wk: The id of the week the date belongs to

-   weekday: The type of the day (Saturday, Sunday, …, Friday).

-   wday: The id of the weekday, starting from Saturday.

-   month: The month of the date.

-   year: The year of the date.

-   event_name_1: If the date includes an event, the name of this event.

-   event_type_1: If the date includes an event, the type of this event.

-   event_name_2: If the date includes a second event, the name of this
    event.

-   event_type_2: If the date includes a second event, the type of this
    event.

-   snap_TX: A binary variable (0 or 1) indicating whether the stores of
    TX, allow SNAP3 purchases on the examined date. 1 indicates that
    SNAP purchases are allowed.

The dataset covers the perido from 29/01/2011 to 19/06/2011. There are
1807 NA values in the event_type_1 columns and 1964 NA values in the
event_type_2 column, indicating that there were no events on the
corresponding dates.

#### Sell Prices

```{r}
# rows and columns
no_of_rows = nrow(sell_prices)
no_of_cols = ncol(sell_prices)
col_names = colnames(sell_prices)

# NA values
rows_na = colSums(is.na(sell_prices))

# Duplicates
dupls = anyDuplicated(sell_prices)


print("Number of rows") 
print(no_of_rows)
print("-----------------------")
print("Number of columns")
print(no_of_cols)
print("-----------------------")
print("Column names:")
print(col_names)
print("-----------------------")

print("Rows with na values")
print(rows_na)
print("-----------------------")
print("Number of duplicates")
print(dupls)
```

#### Sales

split column id to store_id and item_id, so they match with the
corresponding columns from sell_prices and calendar: item_id: item_id,
dept_id, cat_id store_id: state_id, dtate_store_id

Convert dataset to long format. Create column named day_id that matches
with calendar's day_id and column units sold that contains information
about the number of units sold per day.

```{r}
sales_nc <- sales_data %>%
  separate(
    id,
    into = c("item_id", "dept_id", "cat_id", "state_id", "state_store_id", "dataset_name"),
    sep = "_"
  )%>%
  mutate(
    store_id = paste(state_id, state_store_id, sep = "_"),
    item_id = paste(item_id, dept_id, cat_id, sep = "_")  
  ) %>%
  select(-state_id, -state_store_id)
sales <- sales_nc %>%
  pivot_longer(
    cols = starts_with("d_"),
    names_to  = "day_id",
    names_transform = list(day_id = ~ sub("^d_", "id_", .)),
    values_to = "units_sold"
  )
```

```{r}
# rows and columns
no_of_rows = nrow(sales)
no_of_cols = ncol(sales)
col_names = colnames(sales)

# NA values
rows_na = colSums(is.na(sales))

# Duplicates
dupls = anyDuplicated(sales)


print("Number of rows") 
print(no_of_rows)
print("-----------------------")
print("Number of columns")
print(no_of_cols)
print("-----------------------")
print("Column names:")
print(col_names)
print("-----------------------")

print("Rows with na values")
print(rows_na)
print("-----------------------")
print("Number of duplicates")
print(dupls)

summary(sales["units_sold"])
```

The dataset has 1574399 rows. There are no missing values.\
The mean number of units sold was 1.904.

#### **Join calendar and sales information**

A left join keeps all rows from the left table and adds matching columns
from the right table. In this case we use this type of join, because we
want keep all information from sales, but not information from calendar
that are actually useless

```{r}
calendar_sales <- sales%>%
  left_join(calendar, by = "day_id")

head(calendar_sales)
```

#### **Join weekly prices and calendar_sales**

```{r}
full_dataset <- calendar_sales %>%
  left_join(
    sell_prices,
    by = c("store_id", "item_id", "wm_yr_wk")
  )
head(full_dataset)
```

#### **Data Exploration**

Data types.

```{r}
str(full_dataset)
```

```{r}
no_of_rows = nrow(full_dataset)
no_of_cols = ncol(full_dataset)
col_names = colnames(full_dataset)

# NA values
rows_na = colSums(is.na(full_dataset))

# Duplicates
dupls = anyDuplicated(full_dataset)


print("Number of rows") 
print(no_of_rows)
print("-----------------------")
print("Number of columns")
print(no_of_cols)
print("-----------------------")
print("Column names:")
print(col_names)
print("-----------------------")

print("Rows with na values")
print(rows_na)
print("-----------------------")
print("Number of duplicates")
print(dupls)
```

```{r}

# Proportions of missing values for event 1 and event 2 in calendar dataset vs full dataset

mean(is.na(calendar$event_name_1))
mean(is.na(full_dataset$event_name_1))

mean(is.na(calendar$event_name_2))
mean(is.na(full_dataset$event_name_2))
```

```{r}
# Proportion of missing prices at daily level
mean(is.na(full_dataset$sell_price))

# Proportion of missing prices at weekly level
mean(
  is.na(
    full_dataset %>%
      distinct(store_id, item_id, wm_yr_wk, sell_price) %>%
      pull(sell_price)
  )
)

```

The dataset containing all information consists of 1574399 rows, which
is identical to the number of rows in the sales dataset.\
The dataset has 1447657 and 1571106 missing values for event type 1 and
2, respectively. These number are larger than the corresponding numbers
in dataset calendar, however the proportion of missing values in the
datasets match, indicating that the missing event information is
expected and correctly propagated through the join.\
The missing sell_price values occur on weeks where no price was recorded
(e.g., the item was not sold or not available), and these weekly gaps
are repeated across daily observations after joining prices to the sales
data. This is confirmed by the fact that the proportion of missing
values on a daily basis is similar to the proportion of missing values
on a weekly basis.

Zero sales and missing sales

```{r}

full_dataset %>%
  summarise(
    zero_sales = sum(units_sold == 0, na.rm = TRUE),
    na_sales   = sum(is.na(units_sold))
  )
```

Date Continuity

```{r}
full_dataset %>%
  arrange(store_id, item_id, date) %>%
  group_by(store_id, item_id) %>%
  summarise(
    min_date = min(date),
    max_date = max(date),
    n_days   = n(), #number of rows in data 
    span     = as.integer(max_date - min_date) + 1, # number of days with data
    .groups = "drop"
  ) %>%
  filter(n_days != span)
```

There are no missing dates

Summary

```{r}
# TODO: Problem with the notebook, it can't be copied and pasted. 
```

Price per date

```{r}
full_dataset %>%
  filter(!is.na(sell_price)) %>%
  group_by(date) %>%
  summarise(avg_price = mean(sell_price)) %>%
  as_tsibble(index = date) %>%
  autoplot(avg_price)
```

Price - Event

```{r}
full_dataset%>%
  filter(!is.na(sell_price)) %>%
  mutate(event = if_else(is.na(event_name_1), "No event", "Event")) %>%
  group_by(date, event) %>%
  summarise(avg_price = mean(sell_price), .groups = "drop") %>%
  as_tsibble(key = event, index = date) %>%
  autoplot(avg_price)
```

Sales per weekday

```{r}
full_dataset%>%
  group_by(weekday) %>%
  summarise(avg_sales = mean(units_sold)) %>%
  ggplot(aes(weekday, avg_sales)) +
  geom_col()
```

Sales per category

```{r}
full_dataset%>%
  group_by(cat_id) %>%
  summarise(avg_sales = mean(units_sold)) %>%
  ggplot(aes(cat_id, avg_sales)) +
  geom_col()
```

Event Impact

```{r}
full_dataset%>%
  mutate(event = if_else(is.na(event_name_1), "No event", "Event")) %>%
  group_by(date, event) %>%
  summarise(total_sales = sum(units_sold), .groups = "drop") %>%
  as_tsibble(key = event, index = date) %>%
  autoplot(total_sales)
```

SNAP effect

```{r}
 full_dataset%>%
  group_by(date, snap_TX) %>%
  summarise(total_sales = sum(units_sold), .groups = "drop") %>%
  as_tsibble(key = snap_TX, index = date) %>%
  autoplot(total_sales)
```

#### Total Sales

```{r}
full_dataset%>%
  group_by(date) %>%
  summarise(total_sales = sum(units_sold)) %>%
  as_tsibble(index = date) %>%
  autoplot(total_sales)
```

Weekly Sales

```{r}
full_dataset %>%
  group_by(week = yearweek(date)) %>%
  summarise(total_sales = sum(units_sold), .groups = "drop") %>%
  as_tsibble(index = week) %>%
  autoplot(total_sales)
```

STL Decomposition

```{r}
full_dataset %>%
  group_by(week = yearweek(date)) %>%
  summarise(total_sales = sum(units_sold), .groups = "drop") %>%
  as_tsibble(index = week) %>%
  model(STL(total_sales)) %>%
  components() %>%
  autoplot() +
  labs(
    title = "STL Decomposition of Weekly Total Sales",
    x = "Week",
    y = "Units Sold"
  )
```

Increasing trend\
Seasonality (same pattern every year)

```{r}
full_dataset %>%
  group_by(date) %>%
  summarise(total_sales = sum(units_sold), .groups = "drop") %>%
  as_tsibble(index = date) %>%
  model(
    STL(total_sales ~ season(window = "periodic"))
  ) %>%
  components() %>%
  autoplot() +
  labs(
    title = "STL Decomposition of Daily Total Sales",
    x = "Date",
    y = "Units Sold"
  )
```

# Modelling

## Seasonal Naive Model

### 1. Pre-processing

```{r}
# Put all the data into daily time-series format
texas_daily <- full_dataset %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarise(sales = sum(units_sold, na.rm = TRUE), .groups = "drop") %>%
  arrange(date) %>%
  as_tsibble(index = date)

# Determine the horizon and thus the cutoff value
forecast_horizon <- 28
cutoff <- max(texas_daily$date) - days(forecast_horizon)

# Split data into train and validation
ts_train <- texas_daily %>% filter(date <= cutoff)
ts_validation  <- texas_daily %>% filter(date > cutoff)
```

### 2. Fit the model

Fit the seasonal naive model for weekly lag and check the other
baselines for sanity check since the sales data is seasonal, and
seasonal naive should produce the lowest RMSE. We'll use seasonal naive
in the report as a baseline.

```{r}
# Fit the baseline models
models <- ts_train %>%
  model(
    Mean          = MEAN(sales),
    Naive         = NAIVE(sales),
    Seasonal_naive= SNAIVE(sales ~ lag("week")),
    Drift         = RW(sales ~ drift())
  )
```

### 3. Forecast and report the results

```{r}
# Forecast and report 80% & 90% prediction intervals
fc_baseline <- models %>% forecast(h = forecast_horizon, level = c(80, 95))

# Calculate accuracy for the validation predictions
accuracy(fc_baseline, ts_validation)

# Plot the data and predictions
autoplot(ts_train, sales) +
  autolayer(fc_baseline) +
  labs(title = "Baseline Forecasts (28 days)", x = "Date", y = "Sales")
```

### 5. Diagnostics (if needed but I don't think so since these are baselines)

```{r}
# TODO maybe. 
```

## ARIMA

### 1. Pre-processing

No fundamental pre-processing is needed as the dataset that's used in
seasonal naive (i.e., texas_daily) is sufficient for ARIMA modelling.

```{r}
# No fundamental preprocessing is needed compared to dataset that's used
```

Decide whether a (log) transformation is needed or not.

```{r}
# Plot the sales data
autoplot(ts_train, sales)
```

Since the variance is relatively stable across all levels, no
transformation is needed here.

### 2. Fit the model and report the (automatically) chosen model

```{r}
fit_arima <- ts_train %>%
  model(arima = ARIMA(sales))

# report the chosen model
report(fit_arima)
```

### 3. Forecast and report the results

```{r}
# Forecast and report 80% & 90% prediction intervals
fc_arima <- fit_arima %>%
  forecast(h = forecast_horizon, level = c(80, 95))


# Calculate accuracy for the validation predictions
accuracy(fc_arima, ts_validation)

# Plot the predictions
autoplot(ts_train, sales) +
  autolayer(fc_arima) +
  labs(title = "ARIMA Forecast (28 days)", x = "Date", y = "Sales")
```

### 4. Residual Diagnostics

```{r}
# TODO
```
