---
title: "LightGBM"
output: html_notebook
---

In this notebook we create a LightGBM model to forecast the daily sales for the next 28 days.
```{r}
library(data.table)
library(Matrix)
library(dplyr)
library(MLmetrics)
library(lightgbm)
library(stringr)
```

Load Data:
```{r}
set.seed(257)
DATA_DIR <- "/Users/christosgeorghiou/Documents/GitHub/AFCS-M5-Walmart-Sales-Forecasting/data"

PATH_CAL <- file.path(DATA_DIR, "calendar_afcs2025.csv")
PATH_PRICES <- file.path(DATA_DIR, "sell_prices_afcs2025.csv")
PATH_TRAIN <- file.path(DATA_DIR, "sales_train_validation_afcs2025.csv")
PATH_TEST  <- file.path(DATA_DIR, "sales_test_validation_afcs2025.csv")
PATH_SUB   <- file.path(DATA_DIR, "sample_submission_afcs2025.csv")

cal <- fread(PATH_CAL)
prices <- fread(PATH_PRICES)
train_wide <- fread(PATH_TRAIN)
test_wide  <- fread(PATH_TEST)
sub <- fread(PATH_SUB)

names(cal)

```

Pre-processing:
```{r}
d_cols <- grep("^d_", names(train_wide), value = TRUE)

train_long <- melt(
  train_wide,
  id.vars = "id",
  measure.vars = d_cols,
  variable.name = "d",
  value.name = "sales"
)
```

```{r}
calendar <- readr::read_csv(PATH_CAL) %>%
  mutate(
    d = paste0("d_", row_number()),
    date_col = mdy(date)
  ) %>%
  select(
    d, date_col, wm_yr_wk, wday, weekday, month, year,
    event_name_1, event_name_2, snap_TX
  )



```
```{r}
train_long <- train_long %>% select(-any_of("date"))


train_long <- train_long %>%
  left_join(calendar, by = "d") %>%
  rename(date = date_col) %>%
  arrange(id, date)


```

```{r}
names(train_long)
sum(duplicated(names(train_long)))

```
```{r}
train_long <- train_long %>%
  mutate(
    store_id = str_extract(id, "TX_\\d"),
    item_id  = str_extract(id, "FOODS_3_\\d{3}")
  )

```

```{r}
prices <- readr::read_csv(PATH_PRICES)

train_long <- train_long %>%
  left_join(prices, by = c("store_id", "item_id", "wm_yr_wk"))

```
```{r}
summary(train_long$sell_price)

```

Forward-fill prices:
```{r}
dt <- as.data.table(train_long)
setorder(dt, id, date)

dt[, sell_price_ff := nafill(sell_price, type = "locf"), by = id]
```

```{r}
dt[, .(na_rate = mean(is.na(sell_price))), by = id][order(-na_rate)][1:10]


```
Create lag & rolling features:
```{r}
# Lags
for (L in c(1, 7, 14, 28)) {
  dt[, paste0("lag_", L) := shift(sales, n = L, type = "lag"), by = id]
}

# Rolling statistics (past-only)
dt[, roll_mean_7  := frollmean(sales, 7,  align = "right"), by = id]
dt[, roll_mean_28 := frollmean(sales, 28, align = "right"), by = id]
dt[, roll_sd_28   := frollapply(sales, 28, sd, align = "right"), by = id]

# Intermittency
dt[, nz_28 := frollsum(as.integer(sales > 0), 28, align = "right"), by = id]
dt[, zero_ratio_28 := 1 - nz_28 / 28]

```


```{r}
dt[, price_lag_7 := shift(sell_price_ff, 7), by = id]
dt[, price_pct_change_7 :=
      fifelse(price_lag_7 > 0,
              (sell_price_ff - price_lag_7) / price_lag_7,
              0.0)]
dt[, is_price_drop := as.integer(price_pct_change_7 < 0)]

dt[, price_mean_item := mean(sell_price_ff, na.rm = TRUE), by = id]
dt[, price_rel := sell_price_ff / price_mean_item]

```

Time-based validation:
```{r}
max_date <- max(dt$date, na.rm = TRUE)
val_start <- max_date - 27

dt[, target_1 := shift(sales, 1, type = "lead"), by = id]

train_1 <- dt[date < val_start & !is.na(target_1)]
val_1   <- dt[date >= val_start & !is.na(target_1)]

```

```{r}
# Recreate the splits from the cleaned dt
max_date <- max(dt$date, na.rm = TRUE)
val_start <- max_date - 27

dt[, target_1 := shift(sales, 1, type = "lead"), by = id]

train_1 <- dt[date < val_start & !is.na(target_1)]
val_1   <- dt[date >= val_start & !is.na(target_1)]


```

Prepare features:

* wday: Day of the week
* month: Month
* year: Year
* snap_TX: Binary indicator for SNAP eligibility in Texas
* lag_1: Sales from yesterday
* lag_7: Sales from last week
* lag_28: Sales from last month
* roll_mean_7: Average sales over last 7 days
* roll_mean_28: Average sales over last month
* roll_sd_28: Standard deviation of sales over last month
* zero_ratio_28: Fraction of zero-sales days in last 28 days (Distinguishes slow-moving items from consistently selling ones)
* sell_price_ff: Current effective price
* price_rel: Current price relative to item’s average price (Is the item expensive or discounted relative to itself)
* price_pct_change_7: Percentage price change over the last 7 days
* is_price_drop: Binary indicator of a recent price decrease.
```{r}
feat_cols <- c(
  "wday","month","year","snap_TX",
  "lag_1","lag_7","lag_14","lag_28",
  "roll_mean_7","roll_mean_28","roll_sd_28","zero_ratio_28",
  "sell_price_ff","price_rel","price_pct_change_7","is_price_drop"
)


missing_feats <- setdiff(feat_cols, names(dt))
print(missing_feats)
stopifnot(length(missing_feats) == 0)

train_1 <- train_1[complete.cases(train_1[, ..feat_cols])]
val_1   <- val_1[complete.cases(val_1[, ..feat_cols])]

```

```{r}
y_tr <- log1p(train_1$target_1)
y_va <- log1p(val_1$target_1)

X_tr <- as.matrix(train_1[, ..feat_cols])
X_va <- as.matrix(val_1[, ..feat_cols])

```


```{r}
library(lightgbm)

dtrain <- lgb.Dataset(X_tr, label = y_tr)
dval   <- lgb.Dataset(X_va, label = y_va)

params <- list(
  objective = "regression",
  metric = "rmse",
  learning_rate = 0.05,
  num_leaves = 64,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq = 1,
  min_data_in_leaf = 50
)

m1 <- lgb.train(
  params = params,
  data = dtrain,
  nrounds = 5000,
  valids = list(val = dval),
  early_stopping_rounds = 100,
  verbose = 1
)

```


Evaluate:
```{r}
pred <- expm1(predict(m1, X_va))
rmse <- sqrt(mean((val_1$target_1 - pred)^2))
rmse

```
Naive:
```{r}
val_naive <- dt[date >= val_start]
val_naive[, naive := shift(sales, 1, type="lag"), by=id]
rmse_naive <- sqrt(mean((val_naive$target_1 - val_naive$naive)^2, na.rm = TRUE))
rmse_naive

```
Seasonal Naive:
```{r}
val_snaive <- dt[date >= val_start]
val_snaive[, snaive := shift(sales, 7, type="lag"), by=id]
rmse_snaive <- sqrt(mean((val_snaive$target_1 - val_snaive$snaive)^2, na.rm = TRUE))
rmse_snaive

```

